---
title: "MICB425 Portfolio - Microbial Ecological Genomics: From Genomes to Biomes, Spring 2018"
author: "Karl  Abuan, 35575133"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
    toc_float:
      collapsed: no
---
```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::read_chunk("DataScience_AssignmentFeb162018_KarlAbuan_35575133.R")
knitr::read_chunk("DataScience_AssignmentJan262018_KarlAbuan_35575133.R")
```

# Module 01

## Module 01 portfolio check
* Evidence worksheet_01
    + Completion status: X
    + Comments: 
* Evidence worksheet_02
    + Completion status: X
    + Comments:
* Evidence worksheet_03
    + Completion status: X
    + Comments:
* Problem Set_01
    + Completion status: X
    + Comments: 
* Problem Set_02
    + Completion status: X
    + Comments: 
* Writing assessment_01
    + Completion status:
    + Comments: Missing Writing assessment_01 --> See posted screenshots of the essay.
* Additional Readings
    + Completion status:
    + Comments: Need links --> Links are now added. 
    
## Data Science 

* Installation check
    + Completion status: X
    + Comments:
* Portfolio repo setup
    + Completion status: X
    + Comments:
* RMarkdown Pretty PDF Challenge
    + Completion status: X
    + Comments:
* ggplot
    + Completion status: 9/10
    + Comments: Exercise 3 should be at a different taxonomic level than the example plot.

### Installation check
<img src="KA_assignment1.pdf" width="2500" height="2500">

### Portfolio repo setup
git status - is to check is my local repo is updated with the master repo
git fetch then git pull - to pull the files from master repo to local repo
git add . - to place file to the staging message
git commit -m "Commit message" - to commit a file for addtion and to include a message along with it.
git push - to push the file to the master repo

### R Markdown pretty PDF challenge
```{r child = "Pretty_html_Karl_Abuan_35575133.Rmd"}
```

### tidyverse 
```{r DS_AssignmentJan262018}
```

### ggplot
```{r DS_AssignmentFeb162018}
```

## Origins and Earth Systems

### Evidence worksheet 01 "Prokaryotes: The unseen majority"
[Whitman *et al* 1998](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC33863/)
```{r child = "Evidence_worksheet_01_KarlAbuan_35575133.Rmd"}
```

### Evidence worksheet 02 “Life and the Evolution of Earth’s Atmosphere”
Kasting, JF, Siefert, JL. 2002. Life and the Evolution of Earth's Atmosphere. Science. 296:1066-1068. doi: 10.1126/science.1071184.
```{r child = "Evidence_worksheet_02_KarlAbuan_35575133.Rmd"}
```

### Evidence worksheet 03 “The Anthropocene”
Waters, CN, Zalasiewicz, J, Summerhayes, C, Barnosky, AD, Poirier, C, Galuszka, A, Cearreta, A, Edgeworth, M, Ellis, EC, Ellis, M, Jeandel, C, Leinfelder, R, McNeill, JR, Richter, DD, Steffen, W, Syvitski, J, Vidas, D, Wagreich, M, Williams, M, An, ZS, Grinevald, J, Odada, E, Oreskes, N, Wolfe, AP. 2016. The Anthropocene is functionally and stratigraphically distinct from the Holocene. Science. 351:137. doi: 10.1126/science.aad2622.
```{r child = "Evidence_worksheet_03_KarlAbuan_35575133.Rmd"}
```

### Problem set 01 "Prokaryotes: The unseen majority"
[Whitman *et al* 1998](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC33863/)
```{r child = "Problem_set_01_Karl_Abuan_35575133.Rmd"}
```

### Problem set 02 “Microbial Engines”
Falkowski, PG, Fenchel, T, Delong, EF. 2008. The Microbial Engines That Drive Earth's Biogeochemical Cycles. Science. 320:1034-1039. doi: 10.1126/science.1153213
```{r child = "Problem_set_02_Karl_Abuan_35575133.Rmd"}
```

### Writing Assessment 01

### Module 01 references

1. Achenbach, J. 2012. Spaceship Earth: A new view of environmentalism. Washington Post. [Link](https://www.washingtonpost.com/national/health-science/spaceship-earth-a-new-view-of-environmentalism/2011/12/29/gIQAZhH6WP_story.html).

2. Budny, JA. 2017. Book Review: Aerobiology—The Toxicology of Airborne Pathogens and Toxins. International Journal of Toxicology. 36:50-51. doi: 10.1177/1091581816678191. [Link](http://journals.sagepub.com/doi/abs/10.1177/1091581816678191)

3. Canfield, DE, Glazer, AN, Falkowski, PG. 2010. The Evolution and Future of Earth's Nitrogen Cycle. Science. 330:192-196. doi: 10.1126/science.1186120. [20929768](https://www.ncbi.nlm.nih.gov/pubmed/20929768)

4. Falkowski, PG, Fenchel, T, Delong, EF. 2008. The Microbial Engines That Drive Earth's Biogeochemical Cycles. Science. 320:1034-1039. doi: 10.1126/science.1153213. [Link](http://science.sciencemag.org/content/320/5879/1034)

5. Kasting, JF, Siefert, JL. 2002. Life and the Evolution of Earth's Atmosphere. Science. 296:1066-1068. doi: 10.1126/science.1071184. [Link](https://www.ncbi.nlm.nih.gov/pubmed/12004117)

6. Schrag, DP. 2012. Geobiology of the Anthropocene. Fundamentals of Geobiology. 425-436. [Link](https://onlinelibrary.wiley.com/doi/10.1002/9781118280874.ch22)

7. Nisbet, EG, Sleep, NH. 2001. The habitat and nature of early life. Nature. 409:1083-1091. doi: 10.1038/35059210. [Link](https://www.nature.com/articles/35059210)

8. Rockström, J, Steffen, W, Noone, K, Scheffer, M, Teknik- och vetenskapshistoria (bytt namn, 20120201), Skolan för arkitektur och samhällsbyggnad, (ABE), KTH, Filosofi och teknikhistoria. 2009. A safe operating space for humanity. Nature. 461:472-475. doi: 10.1038/461472a. [Link](https://www.nature.com/articles/461472a)

9. Waters, CN, Zalasiewicz, J, Summerhayes, C, Barnosky, AD, Poirier, C, Galuszka, A, Cearreta, A, Edgeworth, M, Ellis, EC, Ellis, M, Jeandel, C, Leinfelder, R, McNeill, JR, Richter, DD, Steffen, W, Syvitski, J, Vidas, D, Wagreich, M, Williams, M, An, ZS, Grinevald, J, Odada, E, Oreskes, N, Wolfe, AP. 2016. The Anthropocene is functionally and stratigraphically distinct from the Holocene. Science. 351:137. doi: 10.1126/science.aad2622. [Link](http://science.sciencemag.org/content/351/6269/aad2622)

10. Whitman WB, Coleman DC, and Wiebe WJ. 1998. Prokaryotes: The unseen majority. Proc Natl Acad Sci USA. 95(12):6578–6583.  [PMC33863](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC33863/)

# Module 02
## Module 02 portfolio check
* Evidence worksheet_04
    + Completion status:
    + Comments: Incomplete --> Now, it's completed.
* Problem Set_03
    + Completion status: X
    + Comments:
* Writing assessment_02
    + CANCELED
* Additional Readings
    + Completion status:
    + Comments: Need links. --> Links are now added. 

### Evidence worksheet 04 “Bacterial Rhodopsin Gene Expression”
Martinez, A, Bradley, AS, Waldbauer, JR, Summons, RE, DeLong, EF. 2007. Proteorhodopsin Photosystem Gene Expression Enables Photophosphorylation in a Heterologous Host. Proc. Natl. Acad. Sci. U. S. A. 104:5590-5595. doi: 10.1073/pnas.0611470104.
```{r child = "Evidence_worksheet_04_KarlAbuan_35575133.Rmd"}
```

### Problem set 03 “Metagenomics: Genomic Analysis of Microbial Communities”
Wooley, JC, Godzik, A, Friedberg, I. 2010. A primer on metagenomics. PLoS Computational Biology. 6:e1000667. doi: 10.1371/journal.pcbi.1000667.
```{r child = "Problem_set_03_Karl_Abuan_35575133.Rmd"}
```

### Module 02 references
1. Madsen, EL. 2005. Opinion: Identifying microorganisms responsible for ecologically significant biogeochemical processes. Nature Reviews Microbiology. 3:439-446. doi: 10.1038/nrmicro1151. [Link](https://www.nature.com/articles/nrmicro1151)

2. Martinez, A, Bradley, AS, Waldbauer, JR, Summons, RE, DeLong, EF. 2007. Proteorhodopsin Photosystem Gene Expression Enables Photophosphorylation in a Heterologous Host. Proc. Natl. Acad. Sci. U. S. A. 104:5590-5595. doi: 10.1073/pnas.0611470104. [17372221](https://www.ncbi.nlm.nih.gov/pubmed/17372221)

3. Rappé, MS, Giovannoni, SJ. 2003. The uncultured microbial majority. Annual Reviews in Microbiology. 57:369-394. doi: 10.1146/annurev.micro.57.030502.090759. [14527284](https://www.ncbi.nlm.nih.gov/pubmed/14527284)

4. Solden, L, Lloyd, K, Wrighton, K. 2016. The bright side of microbial dark matter: lessons learned from the uncultivated majority. Curr. Opin. Microbiol. 31:217-226. doi: 10.1016/j.mib.2016.04.020. [27196505](https://www.ncbi.nlm.nih.gov/pubmed/27196505)

5. Wooley, JC, Godzik, A, Friedberg, I. 2010. A primer on metagenomics. PLoS Computational Biology. 6:e1000667. doi: 10.1371/journal.pcbi.1000667. [20195499](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2829047/)

6. Youssef, NH, Couger, MB, McCully, AL, Criado, AEG, Elshahed, MS. 2015. Assessing the global phylum level diversity within the bacterial domain: A review. Journal of Advanced Research. 6:269-282. doi: 10.1016/j.jare.2014.10.005. [26257925](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4522544/)

# Module 03

## Module 03 portfolio check
* Evidence worksheet_05
    + Completion status: X
    + Comments:
* Problem set_04
    + Completion status: X
    + Comments:
* Writing Assessment_03
    + Completion status:
    + Comments:
* Additional Readings
    + Completion status:
    + Comments: Need links --> Links are now added.
    
### Evidence worksheet 05 “Extensive mosaic stucture”
Welch, RA, Burland, V, Plunkett, G, Redford, P, Roesch, P, Rasko, D, Buckles, EL, Liou, S-, Boutin, A, Hackett, J, Stroud, D, Mayhew, GF, Rose, DJ, Zhou, S, Schwartz, DC, Perna, NT, H. L. T. Mobley, Donnenberg, MS, Blattner, FR. 2002. Extensive Mosaic Structure Revealed by the Complete Genome Sequence of Uropathogenic Escherichia coli. Proc. Natl. Acad. Sci. U. S. A. 99:17020-17024. doi: 10.1073/pnas.252529799.
```{r child = "Evidence_worksheet_05_KarlAbuan_35575133.Rmd"}
```

### Problem set_04 "Counting Candy Microbes"
Kunin, V, Engelbrektson, A, Ochman, H, Hugenholtz, P. 2010. Wrinkles in the rare biosphere: pyrosequencing errors can lead to artificial inflation of diversity estimates. Environ. Microbiol. 12:118-123. doi: 10.1111/j.1462-2920.2009.02051.x.

Sogin, ML, Morrison, HG, Huber, JA, Welch, DM, Huse, SM, Neal, PR, Arrieta, JM, Herndl, GJ. 2006. Microbial Diversity in the Deep Sea and the Underexplored "Rare Biosphere". Proc. Natl. Acad. Sci. U. S. A. 103:12115-12120. doi: 10.1073/pnas.0605127103.
```{r child = "Problem_set_04_Karl_Abuan_35575133.Rmd"}
```

### Writing Assessment 03

<img src="Essays/KA_Module3Essay page 1.pdf" width="2500" height="2500">
<img src="Essays/KA_Module3Essay page 2.pdf" width="2500" height="2500">
<img src="Essays/KA_Module3Essay page 3.pdf" width="2500" height="2500">
<img src="Essays/KA_Module3Essay page 4.pdf" width="2500" height="2500">
<img src="Essays/KA_Module3Essay page 5.pdf" width="2500" height="2500">

### Module 03 references

1. Callahan, BJ, McMurdie, PJ, Holmes, SP. 2017. Exact sequence variants should replace operational taxonomic units in marker-gene data analysis. The ISME Journal. 11:2639. [Link](https://www.ncbi.nlm.nih.gov/pubmed/28731476)

2. Gaudet, AD, Ramer, LM, Nakonechny, J, Cragg, JJ, Ramer, MS. 2010. Small-group learning in an upper-level university biology class enhances academic performance and student attitudes toward group work. PloS One. 5:e15821. doi: 10.1371/journal.pone.0015821. [Link](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0015821)

3. Hallam, SJ, Torres-Beltrán M, Hawley, AK. 2017. Monitoring microbial responses to ocean deoxygenation in a model oxygen minimum zone. Scientific Data. 4:. [Link](https://www.nature.com/articles/sdata2017158)

4. Hawley, AK, Torres-Beltrán M, Zaikova, E, Walsh, DA, Mueller, A, Scofield, M, Kheirandish, S, Payne, C, Pakhomova, L, Bhatia, M. 2017. A compendium of multi-omic sequence information from the Saanich Inlet water column. Scientific Data. 4:170160. [Link](https://www.nature.com/articles/sdata2017160)

5. Kunin, V, Engelbrektson, A, Ochman, H, Hugenholtz, P. 2010. Wrinkles in the rare biosphere: pyrosequencing errors can lead to artificial inflation of diversity estimates. Environ. Microbiol. 12:118-123. doi: 10.1111/j.1462-2920.2009.02051.x. [19725865](https://www.ncbi.nlm.nih.gov/pubmed/19725865)

6. Sogin, ML, Morrison, HG, Huber, JA, Welch, DM, Huse, SM, Neal, PR, Arrieta, JM, Herndl, GJ. 2006. Microbial Diversity in the Deep Sea and the Underexplored "Rare Biosphere". Proc. Natl. Acad. Sci. U. S. A. 103:12115-12120. doi: 10.1073/pnas.0605127103. [16880384](https://www.ncbi.nlm.nih.gov/pubmed/16880384)

7. Torres-Beltrán, M, Hawley, AK, Capelle, D, Zaikova, E, Walsh, DA, Mueller, A, Scofield, M, Payne, C, Pakhomova, L, Kheirandish, S. 2017. A compendium of geochemical information from the Saanich Inlet water column. Scientific Data. 4:170159. [Link](https://www.nature.com/articles/sdata2017159)

8. Welch, RA, Burland, V, Plunkett, G, Redford, P, Roesch, P, Rasko, D, Buckles, EL, Liou, S-, Boutin, A, Hackett, J, Stroud, D, Mayhew, GF, Rose, DJ, Zhou, S, Schwartz, DC, Perna, NT, H. L. T. Mobley, Donnenberg, MS, Blattner, FR. 2002. Extensive Mosaic Structure Revealed by the Complete Genome Sequence of Uropathogenic *Escherichia coli*. Proc. Natl. Acad. Sci. U. S. A. 99:17020-17024. doi: 10.1073/pnas.252529799. [Link](http://www.pnas.org/content/99/26/17020.short)

### Project 1

* CATME account setup and survey
    + Completion status: X
    + Comments: 
* CATME interim group assessment
    + Completion status: X
    + Comments:
* Project 1
    + Report (80%): 
    + Participation (20%):

<img src="Project1/Project1 page 1.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 2.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 3.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 4.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 5.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 6.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 7.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 8.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 9.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 10.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 11.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 12.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 13.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 14.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 15.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 16.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 17.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 18.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 19.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 20.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 21.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 22.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 23.pdf" width="2500" height="2500">
<img src="Project1/Project1 page 24.pdf" width="2500" height="2500">


## Module 04 Portfolio Content
### Project 2

* CATME final group assessment
    + Completion status: 
    + Comments:
* Project 2
    + Report (80%): 
    + Participation (20%):